<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Blue Tits Case study}
-->

Extra-pair paternity in Blue Tits (_Cyanistes caeruleus_): a case study from Westerholz, Bavaria 
========================================================================================================
Supplement to Schlicht, Valcu and Kempenaers _"Spatial patterns of extra-pair paternity: beyond paternity gains and losses"_ (in prep.)
-------------------------------------------------------------------------------------------------------------------------------------


### 1. Getting started
 * Download and install [R](http://cran.rstudio.com/).
 * Open R, and install _expp_ by copying the following line into your R console:  
```{r eval=FALSE }
install.packages("expp")
```  
  
```{r echo=FALSE, message=FALSE}
require(rgeos); require(sp); require(spdep); require(spatstat); require(deldir)
par(mar = c(0,0,1,0) )
```
 * To load _expp_ type:   
```{r } 
require(expp)
```

### 2. Load datasets
For info on the data-sets type: 
```{r eval=FALSE }
help(westerholzBreeding)
help(westerholzEPP)
```
```{r, eval = FALSE}
data(westerholzBreeding)
head(westerholzBreeding)
```
```{r, eval = FALSE}
data(westerholzEPP)
head(westerholzEPP)
```

```{r, echo=FALSE, results='asis'}
data(westerholzBreeding)
knitr::kable(head(westerholzBreeding))
```
***** 
```{r, echo=FALSE, results='asis'}
data(westerholzEPP)
knitr::kable(head(westerholzEPP))
```




### 3. Prepare data

#### 3.1 Split by year (each year needs to be processed separately)

```{r}
b = split(westerholzBreeding, westerholzBreeding$year_)
e = split(westerholzEPP, westerholzEPP$year_) 

# sample sizes by year
lapply(b, nrow)
lapply(e, nrow)
```

#### 3.2 Run a couple of helper functions on both breeding data and extra-pair paternity data 
```{r tidy=FALSE}
breedingDat = lapply(b, SpatialPointsBreeding, coords= ~x+y, id='id', breeding= ~male + female)
eppDat = lapply(e, eppMatrix, pairs = ~ male + female)


```

#### 3.3. Compute Dirichlet polygons based on the `SpatialPointsBreeding` object
```{r}
polygonsDat = lapply(breedingDat, DirichletPolygons)
```  
********************************************************************************

### 4. All the objects are now ready to be processed by the `epp` function.
```{r}
maxlag = 10
O = mapply(FUN = epp, breedingDat, polygonsDat, eppDat, maxlag)
```

```{r results='hide', dpi=100, fig.width=7, fig.height=10, fig.align='left'}
#op = par(mfrow = c(1,2))

for(year in c("2009", "2010") ) { 
  plot(O[[year]], cex = 0.7, lwd = .5, border = "navy" )
  title(main = year)
  }

# par(op)
```
#### Select one nest-box of a given year and zoom in.
```{r, warning=FALSE, dpi=100, fig.width=6, fig.height=8, fig.align='left'}
year = '2010'
box = 110
O10 = O[[year]]
plot(O10 , zoom = box, maxlag = 2,cex = .7,  border = 'white', col = 'grey70', zoom.col = "bisque")

```

```{r results='hide',fig.width=8, fig.height=6}
op = par(mfrow = c(1,2))
    
barplot(O[[1]],relativeValues = TRUE, main = 2009) 
legend(x="topright", legend = c('Observed', 'Potential'), lty = c(1, 2),bty='n')
barplot(O[[2]], relativeValues = TRUE, main = 2010)

par(op)
```

### 5. Fitting a _glmm_ 

#### 5.1 Convert `O` (a list of 2 _epp_ objects) into a `data.frame`.
```{r}
dat = lapply(O, as.data.frame) # a list of data.frame-s
dat = do.call(rbind, dat)
dat$year_ = dat$year__MALE; dat$year__FEMALE = NULL

```

#### 5.2. Data transformations prior to modelling.

##### Rescale rank; rank 1 becames rank 0
```{r}
dat$rank = dat$rank - min(dat$rank)
table(dat$rank)
```
##### Center and re-scale breeding asynchrony (i.e. the difference in laying data between male and female) within each rank.

```{r}
center = function(x) { return(x - mean(x, na.rm = TRUE)) }
scale2 = function(x) { return(x/(2*sd(x, na.rm = TRUE))) }

# Compute asynchrony
dat$asynchrony = abs(dat$layingDate_MALE - dat$layingDate_FEMALE)

#a Compute relative within-rank asynchrony
MALE_splitBy = paste(dat$year_, dat$id_MALE, dat$male, dat$rank, sep = "_")
dat$relative_asynchrony_MALE = unsplit(lapply(split(dat$asynchrony, MALE_splitBy), center), MALE_splitBy)
dat$relative_asynchrony_MALE = scale2(dat$relative_asynchrony_MALE)

FEMALE_splitBy = paste(dat$year_, dat$id_FEMALE, dat$female, dat$rank, sep = "_")
dat$relative_asynchrony_FEMALE = unsplit(lapply(split(dat$asynchrony, FEMALE_splitBy), center), FEMALE_splitBy)
dat$relative_asynchrony_FEMALE = scale2(dat$relative_asynchrony_FEMALE)
```

#### 5.3 Run _glmm_
##### Check if sample size is sufficient for the number of variables we aim to include into the model.
```{r eval=FALSE}
table(dat$epp, dat$year_) #extra-pair frequency by year.
```
```{r echo = FALSE, results='asis'}
knitr::kable(table(dat$epp, dat$year_))
```

##### Run the glmm model (this may take a while depending on your system!).
```{r eval=FALSE}
require(lme4)
dat$age2 = ifelse(dat$male_age_MALE == 'juv', 1, 2)
fm = glmer(epp ~ rank + male_age_MALE + relative_asynchrony_MALE + relative_asynchrony_FEMALE + 
             (1|male) + (1|female) + (1|year_), data = dat, family = binomial)
summary(fm)
```

```
## Generalized linear mixed model fit by maximum likelihood ['glmerMod']
##  Family: binomial ( logit )
## Formula: epp ~ rank + male_age_MALE + relative_asynchrony_MALE + relative_asynchrony_FEMALE +      (1 | male) + (1 | female) + (1 | year_) 
##    Data: dat 
## 
##      AIC      BIC   logLik deviance 
##    588.3    647.5   -286.2    572.3 
## 
## Random effects:
##  Groups Name        Variance Std.Dev.
##  male   (Intercept) 1.12e+00 1.05790 
##  female (Intercept) 9.87e-02 0.31423 
##  year_  (Intercept) 1.54e-05 0.00393 
## Number of obs: 12104, groups: male, 119; female, 117; year_, 2
## 
## Fixed effects:
##                            Estimate Std. Error z value Pr(>|z|)    
## (Intercept)                  -3.275      0.245  -13.39  < 2e-16 ***
## rank                         -1.237      0.156   -7.95  1.9e-15 ***
## male_age_MALEjuv             -1.365      0.449   -3.04   0.0023 ** 
## relative_asynchrony_MALE     -0.386      0.408   -0.95   0.3442    
## relative_asynchrony_FEMALE    0.076      0.376    0.20   0.8397    
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Correlation of Fixed Effects:
##             (Intr) rank   m__MAL r__MAL
## rank        -0.449                     
## ml_g_MALEjv -0.382  0.010              
## rltv_s_MALE  0.091  0.017  0.000       
## rlt__FEMALE -0.008 -0.001 -0.024 -0.345
```









